---
title: "Snowshoe Hares - Exploratory Report"
author: "Julia Wilson"
date: "11/4/2021"
output: 
  html_document: 
    theme: cerulean
    toc: yes
    toc_depth: 5
editor_options: 
  markdown: 
    wrap: 72
---

### Introduction

This report provides an exploratory overview of snowshoe hare
populations from 1998 - 2012 at the Bonanza Creek Experimental Forest in
Alaska (Kielland, K.et al. (2017)). Specifically, the report evaluates
snowshoe hare trap counts over 13 years. It also compares juvenile hare
weights across males and females at three sites; Bonbs, Bonanza Mature,
and Bonanza Riparian. Finally, the report analyzes a
potentialrelationship between hind foot length (mm) and weights (g).

### Data & Analysis

Snowshoe hare trap count, weight and hind foot length data were
collected and made available by Senior Investigator Knut Kielland and
colleagues at the Bonanza Creek (LTER) site in Alaska. Capture -
recapture studies were conducted of snowshoe hares at 5 locales in the
Tanana Valley. The data contain observations for 378 juvenile hares,
collected over 13 years (1998 - 2012). Following exploratory data
visualization, juvenile hare male and female weights are compared by
two-sample t-tests using a significance level (α) f 0.05 throughout.
Differences between the groups are described by Cohen's *d* effect size.
The relationship between weight and hind foot length is explored by a
simple linear regression. All analyses are in R version 4.1.1 using R
studio version 1.4.1717.

```{r echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

# Attach Packages
library(here)
library(tidyverse)
library(janitor)
library(lubridate)
library(kableExtra)
library(effsize)
library(ggbeeswarm)
library(broom)
```

```{r}
# Read in Bonanza Hares data as hare_data
hare_data <- read_csv(here("data", "bonanza_hares.csv"))
```

### Preliminary Results

#### A. Juvenile Hare Trap Counts

```{r}
# Add date column that only adds year 
juvenile_data <- hare_data %>% 
  mutate(date = mdy(date)) %>% 
  mutate(year = year(date)) %>% 
  filter(age == "j") %>% 
  count(year)
```

```{r}
# Data visualization of total number of juvenile hare traps each year
ggplot(data = juvenile_data, 
       aes(x = year, y = n)) +
  geom_bar(stat = "identity", fill = "goldenrod3") + 
  scale_y_continuous(breaks = c(0, 25, 50, 75, 100, 125)) +
  scale_x_continuous(breaks = c(1998, 1999, 2000,2001, 2002, 2003, 2004, 2005, 
                                2006, 2007, 2008, 2009, 2010, 2011,2012)) +
  labs(x = "Year",
       y = "Count (#)",
       title = "Total Juvenile Hare Trap Count by Year (1998 - 2012) (WILSON)",
       caption = "Data Source: Kielland, K.et al. (2017).") + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

**Figure 1: Total Juvenile Hare Trap Count by Year** Total juvenile hare
trap counts per year from 1998 - 2012.

```{r}
# Summary table to analyze takeaways and trends of juvenile hare traps
juvenile_summary <- juvenile_data %>%
  summarize(mean = mean(n, na.rm = TRUE),
            median = median(n, na.rm = TRUE),
            min = min(n, na.rm = TRUE),
            max = max(n, na.rm = TRUE),
            sd = sd(n, na.rm = TRUE)) 
```

##### Juvenile Trap Count Analysis

There is a decrease in juvenile hare trap counts during the time period
1998 - 2012. The minimum count of juvenile hare trap counts reported was
`r juvenile_summary$min` in 2012. The maximum count reported was
`r juvenile_summary$max` in 1999. The overall mean of juvenile hare trap
counts was `r juvenile_summary$mean` and the median was
`r juvenile_summary$median`. The standard deviation was reported at
`r round (juvenile_summary$sd, 1)` hares traped.

To standardize the juvenile hare population in future analyses, the
number of days and traps in the study need to be consistent every year.

#### B. Juvenile Hare Weights

```{r}
# Wrangle the data
juvenile_weight <- hare_data %>% 
  filter(age == "j") %>% 
  mutate(grid = case_when(
    grid == "bonbs" ~ "Bonbs",
    grid == "bonmat" ~ "Bonanza Mature",
    grid == "bonrip" ~ "Bonanza Riparian")) %>% 
  mutate(sex = case_when(
    sex == "f" ~ "Female",
    sex == "m" ~ "Male"))
```

```{r}
# Visualize juvenile weights by site and sex 
ggplot(data = juvenile_weight,
       aes(x = sex, y = weight, color = sex)) +
  geom_beeswarm(aes(show.legend = "none", sex = 3)) +
  facet_wrap(~grid) +
  scale_color_manual(values = c("goldenrod2", "cyan4", "coral")) +
  stat_summary(fun = mean,
               gemo = "point",
               shape = 20,
               size = 1,
               color = "black",
               fill = "black") +
  labs(x = "Sex",
       y = "Weight (g)",
       title = "Juvenile Hare Weight by Sex and Site (WILSON)",
       caption = "Data Source: Kielland, K. et al. (2017).") + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.title = element_text(face = "bold"),
        panel.border = element_rect(color = "black", fill = NA, size = 0.5),
        plot.caption = element_text(hjust = 1))
```

**Figure 2: Juvenile Hare Weight (g) by Sex and Site** Weight(g) for
females, males, and NA at three sites; Bonanza Mature, Bonanza Riparian,
and Bonbs. Gold (Female), Coral (Male), and Grey (NA) points indicate
individual observations for juvenile hare weights (g). The black points
indicate the mean values for each sex, respectively.

**Table 1.** Descriptive statistics (mean, median, max, min, standard
deviation, and sample size) for females and males at three sites; Bonbs,
Bonanza Riparian, and Bonanza Mature. Data: Kielland, K. et al. (2017).

```{r}
# Insert table to show summary values by sex and site
juvenile_weight_summary <- juvenile_weight %>% 
  group_by(grid, sex) %>% 
  drop_na(weight) %>% 
  summarize(mean = mean(weight, na.rm = TRUE),
            median = median(weight, na.rm = TRUE),
            max = max(weight, na.rm = TRUE),
            min = min(weight, na.rm = TRUE),
            sd = sd(weight, na.rm = TRUE),
            sample_size = n()) %>% 
  kable(col.names = c("Site",
                      "Sex",
                      "Mean (g)",
                      "Median (g)",
                      "Max (g)",
                      "Min (g)",
                      "Standard Deviation (g)",
                      "Sample Size"),
        digits = 1) %>% 
 kable_classic() 

juvenile_weight_summary
```

##### Juvenile Weight Analysis

The graph shows a visual difference in counts for juvenile hare weights
for both males and females. Bonanza Riparian has the highest sample size
of juvenile hares at 242. This is followed by Bonanza Mature with a
sample size of 82 and Bonbs at 47. The central tendency for females is
generally lower than the central tendency of males. This is visualized
by the black dots on the graph which represent mean values. The sample
size impacts the visible trends and the accuracy of the data.

#### C. Juvenile Weight Comparison

**Table 2.** Descriptive statistics (mean, median, max, min, standard
deviation, and sample size) for female and male weight (g). Data:
Kielland, K. et al. (2017).

```{r}
# Insert table to visualize summary data by sex
juvenile_weight_summary <- juvenile_weight %>% 
  group_by(sex) %>% 
  summarize(mean = mean(weight, na.rm = TRUE),
            median = median(weight, na.rm = TRUE),
            max = max(weight, na.rm = TRUE),
            min = min(weight, na.rm = TRUE),
            sd = sd(weight, na.rm = TRUE),
            sample_size = n()) %>% 
  kable(col.names = c("Sex",
                      "Mean (g)",
                      "Median (g)",
                      "Max (g)",
                      "Min (g)",
                      "Standard Deviation (g)",
                      "Sample Size"),
        digits = 1) %>% 
  kable_classic()
  

juvenile_weight_summary
```

```{r include = FALSE}
# Create vector for female weights
female_weight <- juvenile_weight %>% 
  filter(sex == "Female") %>% 
  drop_na(weight) %>% 
  pull(weight)

# Create vector for male weights
male_weight <- juvenile_weight %>% 
  filter(sex == "Male") %>% 
  drop_na(weight) %>% 
  pull(weight)

# Plot histograms & qq-plots to check assumptions
hist(female_weight) # Looks normal
qqnorm(female_weight) # Relatively Linear
hist(male_weight) # Irregular distribution
qqnorm(male_weight) # Fairly Linear

# Run t test for female and male weights to find the true difference in means
weight_test <- t.test(male_weight, female_weight)
weight_test

# Get the tidy model results to call outputs in-line: 
weight_test_tidy <- tidy(weight_test)

male_mean <- mean(male_weight)
female_mean <- mean(female_weight)
male_sd <- sd(male_weight)
female_sd <- sd(female_weight)


# Run Cohen test to find effect size 
weight_cohen <- cohen.d(female_weight, male_weight)
weight_cohen
```

##### Juvenile Weight Comparison Analysis

On average, male juvenile hares weight more than female juveniles hares.
The actual difference in mean weights between female and males is
`r round(male_mean, 1) - round(female_mean, 1)` (g), with a small effect
size (Cohen's *d* = `r round(weight_cohen$estimate,2)`). Further
investigation would be needed to identify whether there is a significant
difference between male and female weights despite statistical
significance (t(`r round(weight_test$parameter, 0)`) =
`r round(weight_test$statistic,1)`, p-value:
`r round(weight_test$p.value, 4)`, n = `r count(juvenile_weight)`).

#### D. Juvenile Weight & Hind Foot Length

```{r}
juvenile_foot <- hare_data %>% 
  filter(age == "j")
```

```{r}
# Visualize data comparing juvenile weight & hind foot length
ggplot(data = juvenile_foot, aes(x = hindft, y = weight)) + 
  geom_jitter() +
  labs(x = "Hind Foot Length (mm)",
       y = "Weight (g)",
       title = "Juvenile Weight and Hind Foot Length Comparison (WILSON)",
       caption = "Data Source: Kielland, K. et al. (2017).") +
  geom_smooth(method = "lm", color = "goldenrod2") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.title = element_text(face = "bold"))
```

**Figure 3: Juvenile Weight and Hind Foot Length Comparison**
Relationship between Hind Foot Length (mm) and Weight (g) for juvenile
snowshoe hares recorded from 1998 2012 at Bonanza Creek. Points indicate
individual juvenile hare measurements.

```{r include = FALSE}
# Linear model stored as juvenile_foot
hfoot_weight_lm <- lm(weight ~ hindft, data = juvenile_foot)

# Return complete overview
summary(hfoot_weight_lm)

# Use broom::tidy() function to see the model outputs in a nice format
hfoot_weight_lm_tidy <- broom::tidy(hfoot_weight_lm)
hfoot_weight_lm_glance <- broom::glance(hfoot_weight_lm)

# Find Pearson's r value 
hfoot_weight_cor <- cor.test(juvenile_foot$weight, juvenile_foot$hindft)

# Tidy version of correlation
hfoot_weight_cor_tidy <- tidy(hfoot_weight_cor)

```

##### Juvenile Weight and Hindfoot Analysis

The **slope** is `r round(hfoot_weight_lm$coefficient[2], 2)` (g/mm). On
average we expect weight to increase by
`r round(hfoot_weight_lm$coefficient[2], 2)` for every 1 mm increase in
hind foot length. The **y-intercept** is
`r round(hfoot_weight_lm$coefficient[1], 1)` (g). For a juvenile hare
with a 0 mm hind foot length, we expect an average hare mass of
`r round(hfoot_weight_lm$coefficient[1], 1)` (g).

The **R\^2 value** of `r round(hfoot_weight_lm_glance$r.squared, 1)`
means that 30% of the variance in hare weight is explained by hind foot
length. The **Pearson's *r* Correlation** is
`r round(hfoot_weight_cor$estimate, 1)`. This value represents a
moderate correlation between the change in weight and hind foot length.

Despite the regression line illustrating a positive correlation, the
R\^2 and Pearson's *r* values only support a moderate correlation
between weight and hind foot length.

### Summary

Exploratory data analysis reveals the following initial findings:

-   The total juvenile hare trap counts by year steadily declined from the beginning of sampling (1999) to the end (2012).
-   There is a statistically significant difference in means between male and female juvenile snowshoe hare weights, however the effect size is small.
-   The lack of homoscedasticity in the relationship between hare weight (g) and hind foot length (mm) suggest that the regression model may need to include additional predictor variables to explain the performance of the dependent variable.

Next Steps:

-   Multiple linear regression to understand the relationship between sex, hind foot length and weight

### Citations

David Robinson, Alex Hayes and Simon Couch (2021). broom: Convert
Statistical Objects into Tidy Tibbles. R package version 0.7.9.
<https://CRAN.R-project.org/package=broom>

Erik Clarke and Scott Sherrill-Mix (2017). ggbeeswarm: Categorical
Scatter (Violin Point) Plots. R package version 0.6.0.
[\<https://CRAN.R-project.org/package=ggbeeswarm\>](https://CRAN.R-project.org/package=ggbeeswarm){.uri}.

Garrett Grolemund, Hadley Wickham (2011). Dates and Times Made Easy with
lubridate. Journal of Statistical Software, 40(3), 1-25. URL
[\<https://www.jstatsoft.org/v40/i03/\>](https://www.jstatsoft.org/v40/i03/){.uri}.

Hao Zhu (2021). kableExtra: Construct Complex Table with 'kable' and
Pipe Syntax. R package version 1.3.4.
[\<https://CRAN.R-project.org/package=kableExtra\>](https://CRAN.R-project.org/package=kableExtra){.uri}.

Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER. 2017.
Snowshoe hare physical data in Bonanza Creek Experimental Forest:
1999-Present ver 22. Environmental Data Initiative.
<https://doi.org/10.6073/pasta/03dce4856d79b91557d8e6ce2cbcdc14>

Kirill Müller (2020). here: A Simpler Way to Find Your Files. R package
version 1.0.1. <https://CRAN.R-project.org/package=here>

Sam Firke (2021). janitor: Simple Tools for Examining and Cleaning Dirty
Data. R package version 2.1.0.
<https://CRAN.R-project.org/package=janitor>.

Torchiano M (2020). *effsize: Efficient Effect Size Computation*. doi:
10.5281/zenodo.1480624 (URL: <https://doi.org/10.5281/zenodo.1480624>),
R package version 0.8.1, \<URL:
<https://CRAN.R-project.org/package=effsize>\>.

Wickham et al., (2019). Welcome to the tidyverse. Journal of Open Source
Software, 4(43), 1686,
[\<https://doi.org/10.21105/joss.01686\>](https://doi.org/10.21105/joss.01686){.uri}.
