---
title: "Assignment 4 Task 2"
author: "Julia Wilson"
date: "11/4/2021"
output: 
  html_document: 
    theme: cerulean
    toc: yes
    toc_depth: 5
editor_options: 
  markdown: 
    wrap: 72
---

### Introduction

This report provides an exploratory overview of snowshoe hare
populations from 1998 - 2012 at the Bonanza Creek Experimental Forest in
Alaska (Kielland, K.et al. (2017). Specifically, the report evaluates
snowshoe hare trap counts over 13 years. It also compares juvenile hare
weights across males and females at three sites; Bonrip, Bonanza Mature,
and Bonanza Riparian. Finally, the report analyzes a potential
relationship between hind foot length (mm) and weights (g).

### Data & Analysis

Snowshoe hare trap count, weight and hind foot length data were
collected and made available by Senior Investigator Knut Kielland and
colleagues at the Bonanza Creek (LTER) site in Alaska. The data contain
observations for 378 juvenile hares, collected over 13 years (1998 -
2012). Following exploratory data visualization, juvenile hare male and
female weights are compared by two-sample t-tests using a significance
level (α) f 0.05 throughout. Differences between the groups are
described by Cohen's *d* effect size. The relationship between weight
and hind foot length is explored by a simple linear regression. All
analyses are in R version 4.0.2 using R studio version 1.2.1056.

```{r echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

# Attach Packages
library(here)
library(tidyverse)
library(janitor)
library(lubridate)
library(kableExtra)
library(effsize)
library(ggbeeswarm)
```

```{r}
# Read in Bonanza Hares data as hare_data
hare_data <- read_csv(here("data", "bonanza_hares.csv"))
```

### Preliminary Results
#### A. Juvenile Hare Trap Counts

```{r}
# Add date column that only adds year 
juvenile_data <- hare_data %>% 
  mutate(date = mdy(date)) %>% 
  mutate(year = year(date)) %>% 
  filter(age == "j") %>% 
  count(year)
```

```{r}
# Data visualization of total number of juvenile hare traps each year
ggplot(data = juvenile_data, 
       aes(x = year, y = n)) +
  geom_bar(stat = "identity", fill = "goldenrod3") + 
  scale_y_continuous(breaks = c(0, 25, 50, 75, 100, 125)) +
  scale_x_continuous(breaks = c(1998, 1999, 2000,2001, 2002, 2003, 2004, 2005, 
                                2006, 2007, 2008, 2009, 2010, 2011,2012)) +
  labs(x = "Year",
       y = "Count (#)",
       title = "Total Juvenile Hare Count by Year (1998 - 2012) (WILSON)",
       caption = "Data Source: Kielland, K.et al. (2017).") + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

```

**Figure 1: Total Juvenile Hare Trap Counts by Year** Total juvenile hare trap counts each year from 1998 - 2012.

```{r}
# Summary table to analyze takeaways and trends of juvenile hare traps
juvenile_summary <- juvenile_data %>%
  summarize(mean = mean(n, na.rm = TRUE),
            median = median(n, na.rm = TRUE),
            min = min(n, na.rm = TRUE),
            max = max(n, na.rm = TRUE)) 
```

**Takeaways:** Overall, there is a decrease in juvenile hare trap counts during
the time period 1998 - 2012. The minimum count of juvenile hare trap counts 
reported was `r juvenile_summary$min` in 2012. The maximum count reported was 
`r juvenile_summary$max` in 1999. The overall mean of juvenile hare trap counts 
was `r juvenile_summary$mean` and the median was `r juvenile_summary$median`. 

To standardize the juvenile hare population in future analyses, the number of 
days and number of traps in the study need to be consistent every year.

#### B. Juvenile Hare Weights

```{r}
# Wrangle the data
juvenile_weight <- hare_data %>% 
  filter(age == "j") %>% 
  mutate(grid = case_when(
    grid == "bonbs" ~ "Bonanza Riparian",
    grid == "bonmat" ~ "Bonanza Mature",
    grid == "bonrip" ~ "Bonrip")) %>% 
  mutate(sex = case_when(
    sex == "f" ~ "Female",
    sex == "m" ~ "Male"))
```

```{r}
# Visualize juvenile weights by site and sex 
ggplot(data = juvenile_weight,
       aes(x = sex, y = weight, color = sex)) +
  geom_beeswarm(aes(show.legend = "None"), cex = 3) +
  facet_wrap(~grid) +
  scale_color_manual(values = c("goldenrod2", "cyan4", "coral")) +
  stat_summary(fun = mean,
               gemo = "point",
               shape = 20,
               size = 1,
               color = "black",
               fill = "black") +
  labs(x = "Sex",
       y = "Weight (g)",
       title = "Juvenile Hare Weight by Sex and Site (WILSON)",
       caption = "Data Source: Kielland, K. et al. (2017).") + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.title = element_text(face = "bold"),
        panel.border = element_rect(color = "black", fill = NA, size = 0.5),
        plot.caption = element_text(hjust = 1))
```

**Figure 2: Juvenile Hare Weight (g) by Sex and Site** Compares the
weight (g) of female and male juvenile hares at three sites in Bonanza
Creek; Bonanza Mature, Bonanza Riparian, and Bonrip.

```{r}
# Insert table to show summary values by sex and site
juvenile_weight_summary <- juvenile_weight %>% 
  group_by(grid, sex) %>% 
  summarize(mean = mean(weight, na.rm = TRUE),
            median = median(weight, na.rm = TRUE),
            max = max(weight, na.rm = TRUE),
            min = min(weight, na.rm = TRUE),
            sd = sd(weight, na.rm = TRUE),
            sample_size = n()) %>% 
  kable(caption = "Table 2: Juvenile Weight Summary Statistics",
        col.names = c("Site",
                      "Sex",
                      "Mean (g)",
                      "Median (g)",
                      "Max (g)",
                      "Min (g)",
                      "Standard Deviation (g)",
                      "Sample Size"),
        digits = 1) %>% 
  kable_minimal(html_font = "Georgia", font_size = 12, full_width = FALSE) %>% 
  kable_material(c("striped", "hover", "condensed"))
  

juvenile_weight_summary
```

**Table 1: Juvenile Weight Summary Statistics** This table compares the
mean, median, and sample size of male and female Juvenile Hare weights
(g) at three sites in Bonanza Creek; Bonanza Mature, Bonanza Riparian,
and Bonrip.

**Takeaways:** The graph shows a visual difference in counts for
juvenile hare weights for both males and females. The Bonrip has the
highest count of juvenile hare counts followed by Bonanza Mature and
finally Bonanza Riparian. The central tendency for females is generally
lower than the central tendency of males. The biggest takeaway is that
the number of counts impacts the visible trends and the accuracy of the
data.

#### C. Juvenile Weight Comparison

```{r}
# Insert table to visualize summary data by sex
juvenile_weight_summary <- juvenile_weight %>% 
  group_by(sex) %>% 
  summarize(mean = mean(weight, na.rm = TRUE),
            median = median(weight, na.rm = TRUE),
            max = max(weight, na.rm = TRUE),
            min = min(weight, na.rm = TRUE),
            sd = sd(weight, na.rm = TRUE),
            sample_size = n()) %>% 
  kable(caption = "Table 3: Juvenile Weight Comparison Summary Statistics",
        col.names = c("Sex",
                      "Mean (g)",
                      "Median (g)",
                      "Max (g)",
                      "Min (g)",
                      "Standard Deviation (g)",
                      "Sample Size"),
        digits = 1) %>% 
  kable_minimal(html_font = "Georgia", font_size = 12, full_width = FALSE) %>% 
  kable_material(c("striped", "hover", "condensed"))
  

juvenile_weight_summary
```

**Table 1: Juvenile Weight Comparison Summary Statistics** This table
compares the mean, median, and sample size of male and female juvenile
hare weights (g).

```{r}
# Create vector for female weights
female_weight <- juvenile_weight %>% 
  filter(sex == "Female") %>% 
  drop_na(weight) %>% 
  pull(weight)

# Create vector for male weights
male_weight <- juvenile_weight %>% 
  filter(sex == "Male") %>% 
  drop_na(weight) %>% 
  pull(weight)

# Run t test for female and male weights to find the true difference in means
weight_test <- t.test(male_weight, female_weight)
weight_test

# Run Cohen test to find confidence interval 
weight_cohen <- cohen.d(female_weight, male_weight)
weight_cohen
```

**Analysis:** The actual difference in mean weights between female and
males is 90.5 (g), with a small effect size (Cohen's d =
`r weight_cohen$estimate`). The minimal increase in weight may not
represent a significant difference in weight, despite statistical
significance (t(`r weight_test$parameter`) = `r weight_test$statistic`
p-value: `r weight_test$p.value`, n = `r count(juvenile_weight)`).

**ORIGINAL**

This small effect size indicates a smaller difference between the two
means. The two groups means are separated by 0.29 standard deviations.

However, the t-test has a p-value of `r weight_test$p.value`. This means
that if the samples were drawn from populations with the same mean, the
probability of taking two random samples with means at least as
different as the sample means found by random chance is 0.7%. This takes
our small Cohen's d of `r weight_cohen$estimate` and sample size of 378
into account.

Given the low probability of randomly selecting to random samples with
means at least as different as the sample means found by random chance,
we can reject the null hypothesis.

```{r, include = FALSE}
# Cross reference histogram to explore assumption of t-test
ggplot(data = juvenile_weight, aes(x = weight)) + 
  geom_histogram(color = "red", fill = "white", bins = 7) +
  facet_wrap(~sex) 

ggplot(data = juvenile_weight, aes(sample = weight)) +
  geom_qq() +
  facet_wrap(~sex)
```

#### D. Juvenile Weight & Hind Foot Length

```{r}
juvenile_foot <- hare_data %>% 
  filter(age == "j")
```

```{r}
# Visualize data comparing juvenile weight & hind foot length
ggplot(data = juvenile_foot, aes(x = hindft, y = weight)) + 
  geom_jitter() +
  labs(x = "Hind Foot Length (mm)",
       y = "Weight (g)",
       title = "Juvenile Weight and Hind Foot Length Comparison (WILSON)",
       caption = "Data Source: Kielland, K. et al. (2017).") +
  geom_smooth(method = "lm") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.title = element_text(face = "bold"))
```

**Figure 3: Juvenile Weight and Hind Food Length Comparison** Compares
the relationship between Hind Foot Length (mm) and Weight (g) for
juvenile snowshoe hares recorded from 1998 - 2012 at Bonanza Creek.

```{r}
# Linear model stored as juveinle_foot
hfoot_weight_lm <- lm(weight ~ hindft, data = juvenile_foot)

# Return complete overview
summary(hfoot_weight_lm)

# Use broom::tidy() function to see the model outputs in a nice format
hfoot_weight_tidy <- broom::tidy(hfoot_weight_lm)

hfoot_weight_lm_out <- broom::glance(hfoot_weight_lm)

# Find Pearson's r value 
hfoot_weight_cor <- cor.test(juvenile_foot$weight, juvenile_foot$hindft)
hfoot_weight_cor
```

**Linear Regression Analysis:** The **slope** is
`r round(hfoot_weight_lm$coefficient[2], 2)` (g/mm). On average we
expect weight to increase by
`r round(hfoot_weight_lm$coefficient[2], 2)` for every 1 mm increase in
hind foot length. The **y-intercept** is
`r round(hfoot_weight_lm$coefficient[1], 1)` (g). For a juvenile hare
with a 0 mm hind foot length, we expect an average hare mass of
`r round(hfoot_weight_lm$coefficient[1], 1)` (g).

The **R\^2 value** of `r round(hfoot_weight_lm_out$r.squared, 1)` means
that 30% of the variance in hare weight is explained by hind foot
length. The **Pearson's r Correlation** is
`r round(hfoot_weight_cor$estimate, 1)`. This value represents a
moderate correlation between the change in weight and hind foot length.

Despite the regression line illustrating a positive correlation, the
R\^2 and Pearson's r values only support a moderate correlation between
weight and hind foot length.

### Summary
The total count of juvenile hare traps have declined over
the time period 1998 -- 2012 - Male juvenile hares have a higher mean
weight than female juvenile hares - There is a moderate correlation
between weight and hind foot length

### Citations

Erik Clarke and Scott Sherrill-Mix (2017). ggbeeswarm: Categorical
Scatter (Violin Point) Plots. R package version 0.6.0.
[<https://CRAN.R-project.org/package=ggbeeswarm>](https://CRAN.R-project.org/package=ggbeeswarm){.uri}.

Garrett Grolemund, Hadley Wickham (2011). Dates and Times Made Easy with
lubridate. Journal of Statistical Software, 40(3), 1-25. URL
[<https://www.jstatsoft.org/v40/i03/>](https://www.jstatsoft.org/v40/i03/){.uri}.

Hao Zhu (2021). kableExtra: Construct Complex Table with 'kable' and
Pipe Syntax. R package version 1.3.4.
[<https://CRAN.R-project.org/package=kableExtra>](https://CRAN.R-project.org/package=kableExtra){.uri}.

Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER. 2017.
Snowshoe hare physical data in Bonanza Creek Experimental Forest:
1999-Present ver 22. Environmental Data Initiative.
<https://doi.org/10.6073/pasta/03dce4856d79b91557d8e6ce2cbcdc14>

Kirill Müller (2020). here: A Simpler Way to Find Your Files. R package
version 1.0.1. <https://CRAN.R-project.org/package=here>

Sam Firke (2021). janitor: Simple Tools for Examining and Cleaning Dirty
Data. R package version 2.1.0.
<https://CRAN.R-project.org/package=janitor>.

Torchiano M (2020). *effsize: Efficient Effect Size Computation*. doi:
10.5281/zenodo.1480624 (URL: <https://doi.org/10.5281/zenodo.1480624>),
R package version 0.8.1, \<URL:
<https://CRAN.R-project.org/package=effsize>\>.

Wickham et al., (2019). Welcome to the tidyverse. Journal of Open Source
Software, 4(43), 1686,
[<https://doi.org/10.21105/joss.01686>](https://doi.org/10.21105/joss.01686){.uri}.
